\LARGE \bf Human-Robot Interaction: Tackling the AI Challenges

### 2014-05-07 paper.tex
Typo: Incoming {plan->goal}
Proof-reading section 5 + added TBD notes there

        \node[below=1 of plan] (incoming) {\bf \Large Incoming plan \\ \it Clean the table!};

@@

This last section tries to clarify the ``take home'' message: what are the challenges
that human-robot interaction brings to artificial intelligence, and what is the
current state-of-the-art in this field?

We first restate the challenge of \emph{embodied cognition} and list what we
seen as the main decisional issues that need to be tackled to push forward
human-robot interaction. We then reflect on the importance of knowledge handling
in modern robotics and put our choice of the RDF formalism into perspective.
Finally, we briefly re-discuss two facets of HRI: the importance of
\emph{modelling the humans}, and the difficulty of \emph{acting in presence of
humans}.

@@

Robotics is traditionally regarded as the prototypical instance of \emph{embodied}
artificial intelligence, and this dimension is especially prevalent in
human-robot interaction, where agents have to share a joint physical
environment.

It results in a tight coupling between the symbolic and the geometric realms:
while AI at its origins was mostly a matter of symbolic models, it has been
since recognised that not only the mind is not a purely abstract system,
disconnected from the physical world, but even more, cognition fundamentally
relies on its relation to the physical world (so-called \emph{embodied
cognition}). Varela~\cite{Varela1992} is one of the main discoverer of these
mechanisms, and coined the concept of \emph{enactivism} as the theoretical
framework that study the links between cognition, embodiment and actions.

The challenge of symbol grounding is tightly linked to this issue. It
corresponds to the identification or creation, and then, maintenance of a link
between the symbol (the syntactic form of knowledge the computer will
manipulate) and its semantics, \ie its meaning, anchored in the world (the
relations between the symbol, the referent of the symbol, and mediating minds
is classically referred as the \emph{semantic triangle}, and has been
extensively studied in linguistics). The issue of grounding is well known in
cognitive science and is summarised by Harnard~\cite{Harnad1990} by this
question: ``how the semantic interpretation of a formal symbol system can be
made intrinsic to the system?''. This issue has a major practical importance in
robotic: for a robot to be both endowed with a symbolic representational and
reasoning system, and able to \emph{act} in the physical world, it must ground
its knowledge.

Grounding is implemented at different levels in our architecture. The main source
of grounded symbolic knowledge is the situation assessment module, {\sc Spark}.
Not only it builds symbolic facts from spatial reasoning, but it also relies on
temporal reasoning to track the world state and build explanations to interpret
unexpected perceptions (we mentioned above the example of a disappearing
object).

@@

Grounding also occurs during verbal and non-verbal communication. The {\sc
Dialogs} module grounds new concepts introduced by the speaker by asking questions
until it can attach the new concept to concepts already present in the
knowledge pool (typically, if I ask the robot ``bring me a platypus'', the
robot asks ``what is a platypus''. I may answer ``a mammal'' and the robot will
ask ``what is a mammal?'' and so on until it anchors the new concepts to ones
it already knows). Non verbal interaction (like gestures) is also partially
grounded. When the user points at something, it can say ``give me this'' and
the robot grounds \emph{this} to the pointed artifact.

Because only objects marked with tags are currently recognised (in our typical
lab experiments, it may be a dozen of them), the robot grounding mechanisms work
in a small-sized closed world. This assumption obviously greatly facilitates
the task. It remains to be seen if systematic marking of manipulated objects
(for instance with RFID tags) is a realistic path. It may be the case if the
robot is expected to be used in a well-defined, slowly changing environment.
Real-time recognition of large libraries of objects is another path which may
be viable soon~\cite{Dean2013Fast}.

@@

As a whole, the components that we have introduced so far build a
knowledge-oriented architecture: Knowledge is explicitly stored in one central
and consistent repository of facts, accessible for all modules. It relies on a
\emph{strict formalism} (OWL statements), with a \emph{well defined vocabulary}
(stated in the common-sense ontology). These first two points lead to a
\emph{loosely-coupled architecture} where modules can be removed or replaced
easily by other ones as long as they share the same semantics (modules are
defined by the knowledge they produce).

Also, we adopt a symbolic, reactive, event-driven approach to robot control. By
managing events at the same level as the reasoner, we take full advantage of the
inference abilities of {\sc Oro} to trigger events whose \texttt{true}
conditions can be inferred.

And finally, this architecture allows for the combination of very different
knowledge modalities in a single homogeneous environment, bringing mutual
benefits to components. For instance, the dialogue processing module can
perfectly run without any geometric perception, but its disambiguation routines
can transparently benefit from it when available (since richer symbolic
descriptions of objects are then available).

@@

Knowledge flows in our architecture through a \emph{semantic blackboard}, the
{\sc Oro} server. We rely on Description Logics (OWL) to represent and
manipulate facts.

@@

It also has notable limitations, both fundamental (the suitability of
Description Logics when reasoning on --typically non-monotonic-- commonsense
knowledge is questionable) and practical: RDF triples imply only binary
predicates, which constrains the expressiveness of the system or leads to
cumbersome reifications. Alternatives exist (like {\sc
KnowRob}~\cite{Tenorth2009a}) that mix RDF with more expressive logic languages
like {\sc Prolog} with other limitations, like closed-world reasoning. The
classification performance is another issue: from our experience, with an
ontology sized for a standard experiment (about 100 classes and 200 instances),
classification typically takes about 100ms, which becomes problematic during
interactions.  Besides, the performances are difficult to predict, since a
seemingly inoffensive new statement may indirectly change abruptly the logical
complexity of the whole knowledge model and lead to notable degradation of
classification time.

@@

Alternatives like \emph{Answer Set Programming} have also been
successfully investigated in robotics~\cite{Chen2010,Erdem2012}, in particular
to deal with non-monotonic reasoning. However we did not actually hit any brick wall
while working with OWL ontologies. We may reconsider this choice at a later
stage, but until now it has proven an effective framework to quickly explore
implementations of new cognitive abilities (for instance, it has been
conceptually and technically easy to add support for independent knowledge
models, one per agent the robot interacts with --see section~\ref{sect|tom}).

@@

@@

Perspective taking is this human ability which allows one to put him/herself in
another person's point of view. Studied in psychology
literature~\cite{Flavell1992,Tversky1999}, this ability is crucial when
interacting with people by allowing one to reason on others' understanding of
the world in terms of visual perception, spatial descriptions, affordances and
beliefs, etc.  In the last years these notions have been gradually employed in
HRI.~\cite{Breazeal2006} presents a learning algorithm that takes into account
information about a teacher's visual perspective in order to learn a task.
~\cite{Johnson2005} apply visual perspective taking for action recognition
between two robots.~\cite{Trafton2005} use both visual and spatial perspective
taking to find out the referent indicated by a human partner.

@@

HRI offers new challenges to planning
by adding a complementary constraint: the synthesis of plans that are not only
feasible but legible (in terms of their intention) and compliant with human
preferences. This constraint is often ill-defined, dynamic, context-dependent.

Indeed, we posit that human-robot collaborative task achievement calls for more
subtlety for a robot than autonomous independent task achievement. HATP, by
co-planning for the robot and the human, and modelling social rules as planning
parameters, is a first step in this direction.

\paragraph{Joint actions with humans}\label{sec:soa}
The human presence brings new requirements for robot's abilities both
at the functional and at the deliberative levels~\cite{Klein2004}. The
topics involve motion~\cite{Kulic2007,Berg2004,Madhava2006},
navigation~\cite{Althaus2004,Sisbot2007}, manipulation~\cite{Kemp2007}
in presence of humans as well as perception of human
activities~\cite{Breazeal2001,Burger2008}. Also, when
interacting with humans, robots need to incorporate communication and
collaboration abilities. Several theories dealing with
collaboration~\cite{Cohen1991,Grosz1996,Clark1996} emphasise that
collaborative tasks have specific requirements compared to individual
ones, \eg, since the robot and the person share a common goal, they
have to agree on the manner to realise it, they must show their
commitment to the goal during execution, etc. Several robotic systems
have already been built based on these
theories~\cite{Rich1997,Sidner2005,Breazeal2003} and they
all have shown benefits of this approach. They have also shown how
difficult it is to manage turn-taking between communication partners
and to interleave task realization and communication in a generic
way. Finally, today only few
systems~\cite{Fong2006,Breazeal2003,Sisbot2008} take humans into
account at all levels.

\paragraph{Limits of Disambiguation at Semantic Level}

@@

Another broader direction to head deals with \emph{contexts representation}.
Contexts are currently often limited to the current spatial and temporal
situation. Some projects model offer the possibility to jump in the past or to
switch to another agent's perspective, but in current approaches, selecting a
context always basically consists in retrieving a set of beliefs corresponding
to a situation, and temporarily replacing the current beliefs by those other
ones. This misses the fact that at a given moment, not one but many contexts
co-exist at different scales. We do not want to retrieve one monolithic set of
beliefs, but instead carefully craft a context from several \emph{atomic}
contexts. Techniques for representation of overlapping pools of knowledge
largely remain to be developed, as well as efficient algorithms to retrieve (or
discard) such context-related pools of knowledge. This is a challenge not only
for robotics, but more generally for artificial intelligence.

>>>>

        \node[below=1 of plan] (incoming) {\bf \Large Incoming goal \\ \it Clean the table!};

@@

This last section tries to clarify the ``takeaway'' message: what are the challenges
that human-robot interaction brings to artificial intelligence, and what is the
current state-of-the-art in this field?

We first restate the challenge of \emph{embodied cognition} in general and list
what we see as the main decisional issues that need to be tackled to push
forward human-robot interaction. We then reflect on the importance of knowledge
handling in robotic architectures that deal with complex semantics and put into
perspective our choice of the RDF formalism.  Finally, we briefly re-discuss two
facets of HRI: the importance of \emph{modelling the humans}, and the difficulty
of \emph{acting in presence of humans}.

@@

Robotics is traditionally regarded as the prototypical instance of
\emph{embodied} artificial intelligence, and this dimension is especially
prevalent in human-robot interaction, where agents have to share and collaborate
in a joint physical environment.

This leads to a tight coupling between the symbolic and the geometric realms:
while AI at its origins was mostly a matter of symbolic models, it has been
since recognised that not only the mind is not a purely abstract system,
disconnected from the physical world, but even more, cognition fundamentally
relies on its relation to the physical world (so-called \emph{embodied
cognition}). Varela~\cite{Varela1992} is one of the main discoverer of these
mechanisms, and coined the concept of \emph{enactivism} as the theoretical
framework that study the links between cognition, embodiment and actions.

The challenge of symbol grounding is also tightly linked to this issue. It
corresponds to the identification or creation, and then, maintenance of a link
between the symbol (the syntactic form of knowledge that the computer
manipulates) and its semantics, \ie its meaning, anchored in the world (the
relations between the symbol, the referent of the symbol, and mediating minds
is classically referred as the \emph{semantic triangle} and has been
extensively studied in linguistics). The issue of grounding is well known in
cognitive science and is summarised by Harnard~\cite{Harnad1990} by this
question: ``how the semantic interpretation of a formal symbol system can be
made intrinsic to the system?''. This issue has a major practical importance in
robotic: for a robot to be both endowed with a symbolic representational and
reasoning system, and able to \emph{act} in the physical world, it must ground
its knowledge.

As we have seen, grounding is implemented at different levels in our
architecture. The main source of grounded symbolic knowledge is the situation
assessment module, {\sc Spark}.  Not only it builds symbolic facts from spatial
reasoning, but it also relies on (limited) temporal reasoning to track the world
state and build explanations to interpret unexpected perceptions (like a
suddenly disappearing object).

@@

Grounding also occurs during verbal and non-verbal communication. The {\sc
Dialogs} module grounds new concepts introduced by the speaker by asking questions
until it can attach the new concept to concepts already present in the
knowledge pool (typically, if I ask the robot ``bring me a platypus'', the
robot asks ``what is a platypus''. I may answer ``a mammal'' and the robot will
ask ``what is a mammal?'' and so on until it anchors the new concepts to ones
it already knows). Non verbal interaction (like gestures) is also partially
grounded: We have presented how pointing, for instance, is used by the robot to
ground \emph{``this''} or \emph{``that''} to the pointed artifact.

Because only objects marked with tags are currently recognised (in our typical
lab experiments, it may be a dozen of them), the robot grounding mechanisms work
in a small-sized closed world. This assumption obviously greatly facilitates
the task. It remains to be seen if systematic marking of manipulated objects
(for instance with RFID tags) is a realistic path.\footnote{It may be the case if the
robot is expected to be used in a well-defined, slowly changing environment.
Real-time recognition of large libraries of objects is another path which may
be viable soon~\cite{Dean2013Fast}.}

We see however that we have only scratched the surface of a generic grounding
capability. Context, cultural background, ``naive physics'' knowledge, emotional
state of the human are some of the numerous other determinants that need to be
accounted for when grounding human-robot interaction.

@@

As a whole, the components that we have introduced so far build a
knowledge-oriented architecture: Knowledge is explicitly stored in one central
and consistent repository of facts, accessible for all modules. It relies on a
strict formalism (OWL statements), with a well defined vocabulary
(stated in the common-sense ontology). These first two points lead to a
\emph{loosely-coupled architecture} where modules can be removed or replaced
easily by other ones as long as they share the same semantics (modules are
defined by the knowledge they produce).

Also, we adopt a symbolic, reactive, event-driven approach to robot control. By
managing events at the same level as the reasoner, we take full advantage of the
inference abilities of {\sc Oro} to trigger events whose \texttt{true}
conditions can be (possibly indirectly) inferred.

And finally, this architecture allows for the combination of very different
knowledge modalities in a single homogeneous environment, bringing mutual
benefits to components. For instance, the dialogue processing module can
run without any geometric perception, but its disambiguation routines
can transparently benefit from it when available (since richer symbolic
descriptions of objects are then available).

@@

The {\sc Oro} server relies on Description Logics (OWL) to represent and
manipulate knowledge.

@@

It also has notable limitations, both fundamental (the suitability of
Description Logics when reasoning on --typically non-monotonic-- commonsense
knowledge is questionable) and practical: RDF triples imply only binary
predicates, which constrains the expressiveness of the system or leads to
cumbersome reifications. Alternatives exist (like {\sc
KnowRob}~\cite{Tenorth2009a}) that mix RDF with more expressive logic languages
like {\sc Prolog} with other limitations, like closed-world reasoning.
Classification performance is another issue: from our experience, with an
ontology sized for a standard experiment (about 100 classes and 200 instances),
classification typically takes about 100ms, which becomes problematic during
interactions.  Besides, the performances are difficult to predict, since a
seemingly inoffensive new statement may indirectly change abruptly the logical
complexity of the whole knowledge model and lead to notable degradation of
classification time.

@@

While alternatives like \emph{Answer Set Programming} have been
successfully investigated in robotics~\cite{Chen2010,Erdem2012}, in particular
to deal with non-monotonic reasoning, we did not actually hit any brick wall
while working with OWL ontologies. We may reconsider this choice at a later
stage, but until now it has proven an effective framework to quickly explore
implementations of new cognitive abilities (for instance, it has been
conceptually and technically easy to add support for independent knowledge
models, one per agent the robot interacts with).

@@

\subsubsection{Limits of Disambiguation at Semantic Level}

@@

\fxfatal{Re-organise, rephrase, complement}

@@

HRI offers new challenges to planning by adding a supplementary constraint: the
synthesis of plans that are not only feasible but legible (in terms of their
intention) and compliant with human preferences. This constraint is often
ill-defined, dynamic and context-dependent.  HATP, by co-planning for the robot and
the human, and modelling social rules as planning parameters, is a first step in
this direction.

\fxfatal{What are the next steps?}

\paragraph{Joint actions with humans}\label{sec:soa}
\fxfatal{too many citation -> synthesis needed!}
The human presence brings new requirements for robot's abilities both
at the functional and at the deliberative levels~\cite{Klein2004}. The
topics involve motion~\cite{Kulic2007,Berg2004,Madhava2006},
navigation~\cite{Althaus2004,Sisbot2007}, manipulation~\cite{Kemp2007}
in presence of humans as well as perception of human
activities~\cite{Breazeal2001,Burger2008}. Also, when
interacting with humans, robots need to incorporate communication and
collaboration abilities. Several theories dealing with
collaboration~\cite{Cohen1991,Grosz1996,Clark1996} emphasise that
collaborative tasks have specific requirements compared to individual
ones, \eg, since the robot and the person share a common goal, they
have to agree on the manner to realise it, they must show their
commitment to the goal during execution, etc. Several robotic systems
have already been built based on these
theories~\cite{Rich1997,Sidner2005,Breazeal2003} and they
all have shown benefits of this approach. They have also shown how
difficult it is to manage turn-taking between communication partners
and to interleave task realization and communication in a generic
way. Finally, today only few
systems~\cite{Fong2006,Breazeal2003,Sisbot2008} take humans into
account at all levels.

@@

Another broader direction to head deals with \emph{contexts representation}.
Contexts are currently often limited to the current spatial and temporal
situation. Some models offer the possibility to jump in the past or to
switch to another agent's perspective, but in current approaches, selecting a
context always basically consists in retrieving a set of beliefs corresponding
to a situation, and temporarily replacing the current beliefs by those other
ones. This misses the fact that at a given moment, not one but many contexts
co-exist at different scales. We do not want to retrieve one monolithic set of
beliefs, but instead carefully craft a context from several \emph{atomic}
contexts. Techniques for representation of overlapping pools of knowledge
largely remain to be developed, as well as efficient algorithms to retrieve (or
discard) such context-related pools of knowledge. This is a challenge not only
for robotics, but more generally for artificial intelligence.
